### 2025-11-07 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
Further edits after discussion with G.

\noindent
\textbf{Expected outcomes and significance:}
During the three-year period of the proposed research, we expect to achieve and submit manuscripts on at least the following:
\begin{enumerate}
	\item Demonstrate that the Shannon entropy can be recovered as the maximum entropy increase under mixture for all ensemble spaces, and that the Shannon, Gibbs and von Neumann entropies can be recovered in the specific cases of discrete classical, continuous classical, and quantum mechanics
	\item Show that the geometric structures of classical and quantum mechanics can be recovered from the entropic structure alone 
	\item Demonstrate that information geometry can be generalized to the ensemble space of any physical theory of finite dimensions 
	\item Show that the entropic structure imposes constraints on the space of possible physical theories and rules out some Generalized Probabilistic Theories
\end{enumerate}
\textcolor{blue}{Additional possible outcomes during or subsequent to the period of proposed activities depending on the rate of advancement include [list near-term possible outcomes.  Then discuss possible longer-term future developments stemming from this work]}

\textcolor{blue}{Near-term possible: 
Find what results from information geometry and information theory could be generalized to ensemble spaces. Naturally lends itself to various self-contained contributions by the grad student, undergrads, or individuals outside the project}

\textcolor{blue}{Determine whether the convexity of the mixing entropy can be physically motivated and whether that guarantees a continuous embedding into a locally convex topological vector space.}

\textcolor{blue}{Generalize information geometry to the infinite-dimensional case---would require a locally convex topology to be able to define variations.}

\textcolor{blue}{Determine whether the ensemble spaces can be in general proven to be metrizable and characterize the connection between the metric and the entropy.}

\textcolor{blue}{Generalize classical mechanics and classical field theory.}

\textcolor{blue}{Find a suitable function space that can be used as a model for the case of classical field theories.}

The primary significance of the proposed research is that it will \textcolor{blue}{[has the potential to?] provide a common axiomatic foundation for \emph{all} physical theories.  Such an axiomatic foundation common to all physical theories will serve not only to better ground current theories and elucidate their relationships to one another, but also to constrain potential future physical theories, and perhaps point to assumptions or structures in current theories that hold promise for modification in order to develop new theories.}

\textcolor{blue}{[Rewrite above paragraph along these lines] We don't know where the mathematical structures in physics come from.  Therefore, when people try to generalize them, they do not know in which direction they can be generalized.  What we're doing is instead finding axioms from physical requirements---things the physics must at least give you.  The axiomatics are justified from the physics.  We're doing the due diligence to check that the axioms can be justified from the physics and they're independent of each other.  So when we're doing the generalization, the generalization is better constrained by the physics.}

Regarding the relationship of this work to the present state of knowledge in the field, while there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\cite{hardy2001quantumaxioms,caticha2008,dakic2009,Chiribella2010,Masanes_2011}, we are not aware of any literature \textcolor{blue}{attempting to relate the geometric and entropic structures of quantum mechanics, much less in a general physical theory. One sentence about general approach(es) of those works?  There are people trying to reconstruct QM from physical premises, but these works are typically on finite dimensions and do not generalize to CM, and definitely don't generalize to classical or quantum field theories.}

@@

\noindent
\textbf{Assessment of intellectual merit:} \\
\textcolor{blue}{We will know if the proposed research program is successful if [proofs work!].  [Alternatively, if we conclude that these axioms are insufficient.]}

@@

[Aidala, Uribe, and Carcassi, and once identified, the grad student, will generally meet weekly to discuss the project]

Carcassi will dedicate 20\% of his time to the project and has been the individual most immersed in the technical details of the preliminary work leading to this proposal.  He is (currently part-time) staff in the U-M Department of Physics, a Research Area Specialist Senior, with his salary fully supported through grants.  \textcolor{blue}{He will identify the most promising technical directions, perform a significant fraction of the calculations, serve as the lead writer for publications until the graduate student is at a level to do so, and connect the results of the project to broader results so that the impact of the work is maximized.}  With formal training in computer science and engineering, he will additionally serve as the primary contact to interface with others in these disciplines who may be interested in the work.

Depending on the interests, background, and skills of the Ph.D.~student, there will be different ways for the student to contribute to the research. One possibility is to analyze what constructions and theorems from topological vector spaces can be generalized to topological convex spaces. This will help us better understand the relationship between the two, and possibly give insights into what properties the entropic structure may impose to bridge the gap. Another possibility is to see what results from classical and quantum information theory can be generalized to the entropic structure alone. Yet another possibility, since the JSD is the square of a distance function, is to identify sufficient conditions for a pseudo-distance to be the square of a distance, which may clarify what properties the entropic structure imposes. Developing skills in particular in information theory and information geometry will lead to a wide range of possible next steps in the student's career after their Ph.D., with applications for example in quantum computing and in machine learning across academia and the private sector.

\textcolor{blue}{Mention that we expect to include undergrads}

>>>>

\noindent
\textbf{Expected outcomes and significance:}
During the three-year period of the proposed research, we expect to achieve \textcolor{blue}{and submit manuscripts on} at least the following:
\begin{itemize}
	\item Demonstrate that the Shannon entropy can be recovered as the maximum entropy increase under mixture for all ensemble spaces, and that the Shannon, Gibbs and von Neumann entropies can be recovered in the specific cases of discrete classical, continuous classical, and quantum mechanics.
	\item Show that the geometric structures of classical and quantum mechanics can be recovered from the entropic structure alone.
	\item Demonstrate that information geometry can be generalized to the ensemble space of any physical theory of finite dimensions. 
	\item Show that the entropic structure imposes constraints on the space of possible physical theories and rules out some Generalized Probabilistic Theories.
\end{itemize}
The above expected results should lead naturally to a range of additional possible directions and questions that could be addressed during or subsequent to the grant period, depending on the rate of advancement.  For example:
\begin{itemize}
	\item Identify what results from information geometry and information theory could be generalized to ensemble spaces; this would lend itself to various self-contained contributions by the grad student, undergrads, or individuals outside the project.
	\item Determine whether the convexity of the mixing entropy can be physically motivated and whether that guarantees a continuous embedding into a locally convex topological vector space.
	\item Generalize information geometry to the infinite-dimensional case; this would require a locally convex topology to be able to define variations.
	\item Determine whether the ensemble spaces can be in general proven to be metrizable and characterize the connection between the metric and the entropy.
	\item \textcolor{blue}{Generalize classical mechanics and classical field theory.}
	\item Find a suitable function space that can be used as a model for the case of classical field theories.
\end{itemize}

\textcolor{blue}{The primary significance of the proposed research lies in its approach as well as its specific potential to provide a common axiomatic foundation for \emph{all} physical theories.  Currently within physics, we typically do not know from where the various mathematical structures emerge, making it unclear in which directions they can be generalized in order to develop new theories that are physically meaningful.  Our approach instead seeks to \emph{find axioms from physical requirements}, i.e.~the axiomatics are justified from the physics.  Such an axiomatic foundation common to all physical theories would serve not only to better ground current theories and elucidate their relationships to one another, but also to constrain potential future physical theories}

%We're checking that the axioms are independent of each other.

Regarding the relationship of this work to the present state of knowledge in the field, while there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\cite{hardy2001quantumaxioms,caticha2008,dakic2009,Chiribella2010,Masanes_2011}, we are not aware of any literature \textcolor{blue}{attempting to relate the geometric and entropic structures of quantum mechanics, much less those of a general physical theory. Works trying to reconstruct quantum mechancis from physical premises are typically in finite dimensions, do not generalize to classical mechanics, and do not provide any path to generalize to classical or quantum field theories [add refs or recite some?].}

@@

\noindent
\textbf{Assessment of intellectual merit:} \\
\textcolor{blue}{The proposed research program will be evaluated by the progress made in achieving the stated goals and outcomes.  }

@@

Carcassi will dedicate 20\% of his time to the project and has been the individual most immersed in the technical details of the preliminary work leading to this proposal.  He is (currently part-time) staff in the U-M Department of Physics, a Research Area Specialist Senior, with his salary fully supported through grants.  He, Uribe, and Aidala will together identify the most promising technical directions, he will take the lead in performing a significant fraction of the work, and he will connect the results of the project to broader results so that the impact of the research is maximized.  With formal training in computer science and engineering, he will additionally serve as the primary contact to interface with others in these disciplines who may be interested in the work. Aidala, Uribe, and Carcassi, and once identified, the Ph.D.~student, will generally meet weekly to discuss the project.

Depending on the interests, background, and skills of the Ph.D.~student, there will be different ways for the student to contribute to the research. One possibility is to analyze what constructions and theorems from topological vector spaces can be generalized to topological convex spaces. This will help us better understand the relationship between the two, and possibly give insights into what properties the entropic structure may impose to bridge the gap. Another possibility is to see what results from classical and quantum information theory can be generalized to the entropic structure alone. Yet another possibility, since the JSD is the square of a distance function, is to identify sufficient conditions for a pseudo-distance to be the square of a distance, which may clarify what properties the entropic structure imposes. Developing skills in particular in information theory and information geometry will lead to a wide range of possible next steps in the student's career after their Ph.D., with applications for example in quantum computing and in machine learning across academia and the private sector.  We additionally anticipate involving one or more undergraduates in the project, identifying self-contained pieces of the project to which they could contribute.  \textcolor{blue}{For example, starting from a few specific results from information geometry or information theory, they could work on trying generalize these results to ensemble spaces.}
